---
title: Notebook 2 Cluster
author: Erwin Lares
date: '2026-07-01'
categories:
- data-science
- R
- linguistics
- pottery
- aikido
- brug
- personal
- productivity
- coding
- Quarto
- R
- reproducible research
- research workflow
- command line
- Rscript
- CHTC
- portable code
- data subsetting
- toolero
description: |
  A practical guide to starting a research coding project with a structure that can grow from local exploration to reproducible analysis, containerized execution, and high-throughput computing. Part 0 of the "From the Notebook to the Cluster" series.
engine: markdown
---


## The long game

In Part 0, we gave the project a home. We created a project structure that was not only useful for local work, but also prepared for the later steps in this series: writing a reproducible analysis document, deriving a runnable script, containerizing the software environment, and eventually submitting the work to CHTC.

That order matters.

A project that starts with a stable structure is easier to reason about. Raw data has a predictable location. Results have a predictable destination. Future job inputs have a place to go. The analysis is no longer just a script somewhere on a laptop. It belongs to a project.

In Part 1, we create the document that will carry the analysis.

The point of this post is not simply that Quarto is a good reporting tool, although it is. The larger point is that a Quarto document can become the source of truth for an analysis. Prose, code, outputs, and decisions live together. From that document, we can derive the standalone `.R` script that later stages of the workflow will need.

That distinction is important for the long game:

```text
.qmd document  -> human-readable source of truth
.R script      -> machine-executable derived artifact
```

The `.qmd` is where we explain the analysis. The `.R` file is what we will eventually run outside the document, containerize, and hand to a scheduler.

If those two files are maintained separately by hand, they can drift apart. If the `.R` file is derived from the `.qmd`, the relationship is clearer: edit the document, render the document, regenerate the script.

That is the role of `toolero::create_qmd()` in this workflow.

## Where this post sits in the sequence

This series follows a deliberate path:

```text
Part 0  create the project home     -> toolero::init_project()
Part 1  create the document/code    -> toolero::create_qmd()
Part 2  containerize the project    -> containr
Part 3  submit the first job        -> submitr single-job workflow
Part 4  submit many jobs            -> toolero::write_by_group() + submitr
```

Each step prepares for the next one.

In Part 0, we used `toolero::init_project()` to create a project called `palmer-penguins-analysis`. We also made one project-specific choice: because we know this analysis will eventually be split into job-sized inputs, we created `data/jobs/` from the beginning.

At the start of this post, assume the project has this shape:

```text
palmer-penguins-analysis/
├── data/
│   └── jobs/
├── data-raw/
│   └── palmer-penguins.csv
├── docs/
├── images/
├── plots/
├── R/
├── results/
└── scripts/
```

The dataset lives at:

```text
data-raw/palmer-penguins.csv
```

That path will appear throughout the post. The name is intentionally specific. A file called `data.csv` is convenient in the moment, but it becomes vague quickly. A file called `palmer-penguins.csv` tells us what the input is before we open it.

## Why start with the document?

Most research projects start as a script. That is understandable. A script is direct. It lets you load the data, try an idea, make a plot, and keep moving.

The difficulty appears later.

A script can tell you what the code did, but it often does not tell you why the analysis was written that way. It may not record which decisions were exploratory and which became part of the final workflow. It may not explain why a particular input path, model, filter, or figure exists. When the project needs to be shared, reviewed, revisited, or moved to another machine, that missing context becomes expensive.

A Quarto document helps because it keeps the reasoning close to the code. The explanation, the executable code, and the outputs live in one place. The analysis is no longer a script plus a separate set of notes plus a folder of copied figures. It is a document that can be read, rendered, and inspected.

There is also a practical benefit that is easy to underestimate: consolidation. A common research workflow spreads the work across several tools: code in an editor, tables and charts in a spreadsheet, narrative in a document editor, and a manual copy-paste step connecting them all. Every time the data changes, the cycle repeats.

Quarto collapses much of that work into a single source. Tables and visualizations are generated directly from the code and appear in the document automatically. When the data changes, a render updates the outputs. There is no figure to re-export by hand and no table to manually reconcile.

This does not mean that every project should remain a single document forever. That would be too strong a claim. As a project grows, reusable functions may belong in `R/`, helper scripts may belong in `scripts/`, and long-running computation may need to move elsewhere. But as a starting point, the Quarto document gives the project a narrative spine.

The goal is not purity. The goal is to keep the analysis understandable, reproducible, and easy to move.

## Create the document with toolero

The `toolero` package includes `create_qmd()`, a function that scaffolds a Quarto analysis document with reproducible project conventions in mind.

First, load `toolero`:

```r
library(toolero)
```

From inside the project root, create the analysis document:

```r
create_qmd(
  path     = ".",
  filename = "notebook2cluster-p1-start-with-the-document.qmd"
)
```

This does more than create an empty `.qmd` file. It gives us a starting document and, by default, sets up the mechanism that helps prevent code drift: a post-render `purl()` workflow that extracts the R code from the document into a companion `.R` script.

The important relationship is one-way:

```text
edit the .qmd  -> render the document  -> derive the .R script
```

The `.qmd` is the file we maintain. The `.R` script is derived from it.

That is the design choice that matters. If we later run the script from the command line, inside a container, or as part of a CHTC job, we want confidence that the script reflects the current document. We do not want two parallel versions of the analysis that have to be kept in sync by memory.

## Use YAML to make repeated setup less repetitive

A Quarto document begins with YAML metadata. At its simplest, YAML records the title, format, and other rendering options for the document. In practice, it often becomes a place where repeated setup accumulates: author name, affiliation, table of contents settings, theme, CSS, project-specific parameters, and other conventions.

This is one of the reasons `create_qmd()` is useful. If you regularly create documents with the same author information, institutional metadata, preferred options, or parameters, you should not have to rebuild those fields by hand every time.

`create_qmd()` can use a YAML configuration file to pre-populate the document header:

```r
create_qmd(
  path      = ".",
  filename  = "notebook2cluster-p1-start-with-the-document.qmd",
  yaml_data = "path/to/my-config.yml"
)
```

For this post, the key parameter is the input data file:

```yaml
params:
  data_file: "data-raw/palmer-penguins.csv"
```

That parameter gives the document a default input when it is rendered with Quarto. Later, when the same code is run with `Rscript`, the input can be supplied as a command-line argument instead.

This is the beginning of portability. The same analysis can be run in more than one context without editing the code each time.

## Before we write the analysis

Before writing analysis code, it is worth naming the constraints we want the document to satisfy.

The document should:

* read the original dataset from `data-raw/palmer-penguins.csv`;
* write tabular outputs to `results/`;
* write visual outputs to `plots/`;
* behave correctly when rendered as a Quarto document;
* behave correctly when run as a standalone `.R` script;
* make the later transition to containerized and scheduled execution easier.

Those are not advanced requirements. They are the normal requirements of a project that may eventually leave your laptop.

The safest time to account for them is now.

## A minimal analysis thread

The analysis in this post is intentionally small. We will use the Palmer Penguins dataset to calculate summary statistics for flipper length and visualize the distribution of flipper length by species.

The analysis itself is not the point. The pattern is the point.

A small analysis lets us focus on the structure:

```{mermaid}
%%| label: fig-mermaid
%%| fig-cap: "Analysis workflow"
flowchart LR
  A[Load libraries<br>and data] --> B[Prepare data]
  B --> C[Calculate summaries]
  C --> D[Render table]
  C --> E[Render plot]
  D --> F[Save outputs]
  E --> F
```

@fig-mermaid gives the bird's-eye view. We load the data, compute summaries, render outputs, and save the deliverables in predictable locations.

That final step matters. A result printed to the screen may be useful during exploration, but a reproducible workflow needs files that can be inspected, shared, archived, or retrieved after a job runs somewhere else.

## Resolve inputs across execution contexts

One of the subtle challenges in this workflow is avoiding code drift across execution contexts.

It is very easy to begin with one version of the logic for interactive work in RStudio, introduce a slightly different version for `quarto render`, and eventually add another version for `Rscript` when the analysis needs to run headlessly. Over time, those entry points can drift apart. A path gets updated in one place but not another. A parameter is handled differently. Behavior that worked locally no longer matches the rendered or scheduled version.

The result is duplicated logic and a project that becomes harder to trust.

The code below keeps the input resolution in one place. In an interactive session, the data path is convenient for local development. In a Quarto render, the path comes from the YAML `params` block. In an `Rscript` call, the path comes from the command-line argument, which is the pattern a job scheduler can use later.

```{r}
library(tidyverse)
library(gt)
library(toolero)

context <- detect_execution_context()

args <- commandArgs(trailingOnly = TRUE)

data_file <- switch(context,
  interactive = "data-raw/palmer-penguins.csv",
  quarto      = params$data_file,
  rscript     = args[[1]]
)

is_rscript <- identical(context, "rscript")
should_render_outputs <- !is_rscript

data <- readr::read_csv(data_file)
```

The important thing is that `data_file` is resolved once. The rest of the analysis does not need to know whether the code was run interactively, rendered as a document, or called from the command line.

That is a small design decision, but it has a long reach. It is one of the things that makes the `.qmd` and the derived `.R` script portable.

## Tabular results

We can now calculate a small set of summary statistics for flipper length across the three penguin species.

The code below groups the data by species and computes the minimum, maximum, mean, and standard deviation of `flipper_length_mm`. The resulting data frame is then formatted as a table with `gt` when the document is rendered or run interactively.

```{r}
#| label: tbl-flipper-summary
#| tbl-cap: "Summary statistics for flipper length by species"
flipper_stats <- data |>
  group_by(species) |>
  summarize(
    min_flipper_length  = min(flipper_length_mm, na.rm = TRUE),
    max_flipper_length  = max(flipper_length_mm, na.rm = TRUE),
    mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
    sd_flipper_length   = sd(flipper_length_mm, na.rm = TRUE),
    .groups = "drop"
  )

flipper_stats_tbl <- flipper_stats |>
  gt() |>
  fmt_number(decimals = 0) |>
  tab_header(
    title = "Flipper Length by Species",
    subtitle = "Summary statistics in millimeters"
  )

if (should_render_outputs) flipper_stats_tbl
```

@tbl-flipper-summary shows a familiar pattern in the Palmer Penguins data: Gentoo penguins have longer flippers than Adelie and Chinstrap penguins. The table is useful because it makes the numerical pattern explicit.

For the long game, there is another point worth noticing. The object `flipper_stats` is not only something we can print in the document. It is also something we can save as a file.

That distinction matters when the analysis runs outside the document.

## Visualizing distributions

A table summarizes the pattern, but a plot makes the distribution easier to inspect.

The code below visualizes flipper length across species using three complementary layers: a violin plot for the distribution, a boxplot for the summary, and jittered points for the individual observations.

```{r}
#| label: fig-flipper-distribution
#| fig-cap: "Distribution of flipper length by species"
p <- data |>
  ggplot(aes(
    x = species,
    y = flipper_length_mm,
    color = species,
    fill = species
  )) +
  geom_violin(alpha = 0.15, linewidth = 0) +
  geom_boxplot(alpha = 0.4, width = 0.3, outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.5, size = 1.5) +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title    = "Flipper Length by Species",
    subtitle = "Distribution of flipper length in millimeters",
    x        = NULL,
    y        = "Flipper length (mm)",
    caption  = "Data: Palmer Penguins"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

if (should_render_outputs) p
```

@fig-flipper-distribution shows the same general result as the table, but with more texture. Adelie and Chinstrap penguins overlap more strongly, while Gentoo penguins are visibly larger by this measure. The point is not that this is a complicated analysis. The point is that the document lets us keep the explanation, code, table, and plot together.

## Save analysis outputs

In many analyses, the deliverables are not only objects printed to the screen. They are files saved for later use: summary tables, plots, model objects, logs, or other artifacts that a collaborator, manuscript, or later workflow will need.

For this project, the folder conventions were established in Part 0:

```text
results/   # tabular or non-visual outputs
plots/     # visual outputs
```

The code below saves the summary statistics to `results/` and the plot to `plots/`.

```{r}
results_dir <- "results"
plots_dir <- "plots"

for (dir in c(results_dir, plots_dir)) {
  if (!dir.exists(dir)) {
    cli::cli_alert_info("Creating directory: {.path {dir}}")

    dir_created <- dir.create(dir, recursive = TRUE)

    if (!dir_created) {
      cli::cli_abort("Could not create directory: {.path {dir}}")
    }
  } else {
    cli::cli_alert_info("Using existing directory: {.path {dir}}")
  }
}

csv_file <- file.path(results_dir, "flipper-stats.csv")
png_file <- file.path(plots_dir, "flipper-length-by-species.png")

cli::cli_alert_info("Writing summary statistics to {.path {csv_file}}")
readr::write_csv(flipper_stats, csv_file)

cli::cli_alert_info("Saving plot to {.path {png_file}}")
ggplot2::ggsave(
  filename = png_file,
  plot     = p,
  width    = 7,
  height   = 5,
  dpi      = 300
)

cli::cli_alert_success("Results saved successfully.")
```

This is a modest step, but it is a useful habit. The script does not assume that a person will manually create output folders before running it. It checks for the directories it needs and writes deliverables to predictable locations.

That matters later, when the same code runs without anyone watching it.

## Derive the runnable R script

When the analysis needs to run at scale, the document-rendering overhead may be unnecessary. At that point, we want the code, not the narrative.

This is where `create_qmd()` does more than create a blank Quarto document.

By default, `create_qmd()` can set up the project so that rendering the `.qmd` also derives a standalone `.R` script from the document. That behavior is controlled by the `use_purl` argument.

```r
create_qmd(
  path     = ".",
  filename = "notebook2cluster-p1-start-with-the-document.qmd",
  use_purl = TRUE
)
```

With `use_purl = TRUE`, `toolero` creates the pieces needed for an automatic post-render workflow. The project gets a `_quarto.yml` file that registers a post-render hook, and that hook runs a `purl.R` script after the document renders.

Conceptually, the workflow looks like this:

```text
render the .qmd
  -> Quarto reads _quarto.yml
      -> post-render runs purl.R
          -> purl.R derives the companion .R script
```

The result is the same relationship we want for the long game:

```text
notebook2cluster-p1-start-with-the-document.qmd
  -> notebook2cluster-p1-start-with-the-document.R
```

You maintain the `.qmd`. The `.R` script is regenerated from it.

Under the hood, the post-render script relies on `knitr::purl()`, which extracts R code chunks from a literate document and writes them to a standalone script. The source template for that helper script lives in the `toolero` package at `inst/templates/purl.R`. When `create_qmd(use_purl = TRUE)` is called, `toolero` copies or writes the needed post-render helper into the project so the project can regenerate its own `.R` script when rendered.

The exact implementation detail is less important than the direction of authority:

```text
.qmd is maintained by the researcher
.R is derived by the project machinery
```

That direction is what prevents code drift. If the document and the script are both edited by hand, they can become two similar but different analyses. If the script is derived from the document, the relationship is clearer: edit the document, render the document, regenerate the script.

The `.qmd` is the source of truth. The `.R` file is the derived artifact.

## Prepare for job-sized inputs

At this point, we have a document that can render an analysis and save outputs. That is enough for a local workflow. It is not yet enough for a high-throughput workflow.

HTC becomes useful when a larger task can be split into many independent pieces. For this toy example, splitting by penguin species is not computationally necessary. It is useful because it illustrates the pattern we will need later: one input file per job.

In Part 0, we created `data/jobs/` for this future purpose. Now we can use it.

`toolero::write_by_group()` splits a data frame by a grouping column, writes one CSV file per group, and optionally writes a manifest that records the files created.

```{r}
toolero::write_by_group(
  data,
  group_col  = "species",
  output_dir = "data/jobs",
  manifest   = TRUE
)
```

This gives the future submission workflow a set of job-sized inputs. Conceptually, the structure is now:

```text
data-raw/palmer-penguins.csv   # original input data
data/jobs/                     # split inputs for later jobs
```

The raw data remains untouched. The job inputs are derived from it.

That distinction is worth preserving.

## Run the derived script from the command line

Once the `.R` script has been generated and the data has been split into subsets, each subset can be processed independently by passing it as a command-line argument.

For example:

```bash
Rscript notebook2cluster-p1-start-with-the-document.R data/jobs/adelie.csv
```

This is the same kind of call that a job scheduler can use later. At small scale, you run one command manually. At larger scale, a submit file runs the same script many times, each time with a different input file.

The fact that the same script can run in both situations is the point. The portability was built in earlier, when we resolved the input file based on the execution context.

When the time comes to scale, the transition should be a change in infrastructure, not a rewrite of the analysis.

## What we have after Part 1

At the end of Part 1, the project has moved from a structured home to a structured analysis.

We now have:

* a Quarto document that explains the analysis;
* a derived `.R` script that can run outside the document;
* an input path that works across interactive, Quarto, and `Rscript` contexts;
* a summary table saved under `results/`;
* a plot saved under `plots/`;
* job-sized input files saved under `data/jobs/`;
* a pattern that can be containerized and eventually submitted to CHTC.

The important thing is not that the Palmer Penguins analysis is complex. It is not. The important thing is that the project now has a shape that can move.

## What comes next

In Part 2, we will turn to the software environment.

Running a script outside the document is one step. Running it somewhere other than your laptop is another. For that, the project needs a portable software environment: the right R version, the right package versions, and the system libraries those packages require.

That is the role of `containr`.

The project already has the pieces that make containerization easier: a stable project structure, a clear input path, a derived script, saved outputs, and an `renv.lock` file. In the next part, we will use those pieces to build a container image.

For now, the document is doing what we need it to do.

It explains the analysis for humans and produces a script that machines can run.



